---
title: "Assignment 3 - Diamond Price Estimation"
author: "Umut Mete Saka"
output: 
 html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)
```

## Introduction

This Assignment aims to imply explanatory data analysis on diamonds data from ggplot2 package and build a multiple linear regression model on the data to predict the price indicator using the other inputs. This is going to be a supervised learning model. 

## Starting Points

First of all I am going to load required packages. My data is in the ggplot2 package which is in tidyverse. Therefore I do not need to import my data exclusively.

```{r}
library(tidyverse)
library(lubridate)
library(tidyr)
library(knitr)
library(gridExtra)
library(tree)
```

Now I can glimpse my data to understand its characteristics.

```{r}
glimpse(diamonds)
```

There are 10 attributes in the data. These attributes are:


- Carat : Weight of the diamond 

- Cut : Quality of the cut 

- Color : Diamond colour, from D (best) to J (worst)

- Clarity : A measurement of how clear the diamond is (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best))

- Depth : Total depth percentage = z / mean(x, y) 

- Table : Width of top of diamond relative to widest point

- Price : Price in US dollars  *Which is going to be my output in the predictive model* 

- x : Length in mm 

- y : Width in mm

- z : Depth in mm 

I observe that there are 6 numerical input attributes. So I am going to see the correlation between these attributes and price.

```{r}
Diamonds_num <-  diamonds %>%
  select(carat,depth,table,price,x,y,z)

kable(cor(Diamonds_num$price, Diamonds_num), caption = "Correlation between price")


```


From the correlation table I can see that, carat, x, y and z have high positive correlation with price. However table and depth seems uncorrelated. So I am going to plot the correlated attributes with price. 

```{r}
plot1 <- Diamonds_num %>% select(-table, -depth) %>% filter(x>3)%>% ggplot(aes(price,x)) + geom_line(color = "blue") 
plot2 <- Diamonds_num %>% select(-table, -depth) %>% filter(y<20) %>% ggplot(aes(price,y)) + geom_line(color = "red") 
plot3 <- Diamonds_num %>% select(-table, -depth) %>% filter(z<10) %>% ggplot(aes(price,z)) + geom_line(color = "green") 
plot4 <- Diamonds_num %>% select(-table, -depth) %>% ggplot(aes(price,carat)) + geom_line(color = "orange") 

grid.arrange(plot1,plot2,plot3,plot4,nrow = 2)
```

Now I want to see the effect of the categorical variables on price. There are 3 categorical variables which are cut, color and clarity. 

```{r}

Diamonds_cat <-  diamonds %>% select(price,cut,color,clarity)

Diamonds_cat %>% select(price,cut)  %>% group_by(cut) %>% mutate(mean_price = mean(price)) %>% ggplot(aes(x = cut)) + geom_point(aes(y = mean_price),color = "blue",cex= 1.3) + geom_point(aes(y=mean(price)),color = "red",cex = 1) +xlab("Cut")+ylab("Price")+ggtitle("Price vs Cut")
```
 It seems that some cut types has significant effect on the price. Premium and Fair cuts are much more expensive than average price and Ideal cut is much lower.
 
```{r}


Diamonds_cat %>% select(price,color) %>% group_by(color) %>% mutate(mean_price = mean(price)) %>% ggplot(aes(x = color)) + geom_point(aes(y = mean_price),color = "blue",cex= 1.3) + geom_point(aes(y=mean(price)),color = "red",cex = 1) +xlab("Color")+ylab("Price")+ggtitle("Price vs Color")

```
 
Also Color seems to have significant effect on Price.

 
```{r}


Diamonds_cat %>% select(price,clarity) %>% group_by(clarity) %>% mutate(mean_price = mean(price)) %>% ggplot(aes(x = clarity)) + geom_point(aes(y = mean_price),color = "blue",cex= 1.3) + geom_point(aes(y=mean(price)),color = "red",cex = 1) +xlab("Clarity")+ylab("Price")+ggtitle("Price vs Clarity")

```

Also some types of clarity has effect on price. 


## Predictive Models

First I will start with dividing data into test and train datasets. Train data set will use random 70% of the data.

```{r}
set.seed(48)
a <- sample(1:nrow(diamonds),nrow(diamonds)*0.7)
diamonds_train <- diamonds[a,]
diamonds_test <- diamonds[-a,]
```

#### Multiple Linear Regression
I am going to build a multiple linear regression model to predict the price. In my first model I am going to use every input attribute and see their significance levels.

```{r}
model1 <- lm(price~.,diamonds_train)

summary(model1)
```

My r^2 value is 0.92 which shows a nice fit to the data.

Now I will run the model on test data and calculate Mean Absolute Percentage Error. 

```{r}
diamonds_test <-  diamonds_test %>% mutate(predicted_values_lm = predict(model1,diamonds_test))

MAPE_model_lm <- mean(abs((diamonds_test$price - diamonds_test$predicted_values_lm) / diamonds_test$price)) * 100

```
 My Multiple Linear Regression model worked with `r signif(MAPE_model_lm,digits = 4)`% Error.
 

#### Tree Model

My second model is a decision tree.
```{r}
model2 <- tree(price~.,diamonds_train)
summary(model2)
```

It is a tree with 6 nodes. Lets visualize it:

```{r}
plot(model2)
text(model2)
```


```{r}


diamonds_test <-  diamonds_test %>% mutate(predicted_values_tree = predict(model2,diamonds_test))

MAPE_model_tree <- mean(abs((diamonds_test$price - diamonds_test$predicted_values_tree) / diamonds_test$price)) * 100

```

 
My tree model worked with `r signif(MAPE_model_tree,digits = 4)`% Error.
 

Finally I want to visualize the model outcomes with the real data.

```{r}
b <- seq(1:nrow(diamonds_test))
diamonds_test %>% mutate(index=b) %>% ggplot(aes(x=price)) + geom_point(aes(y=predicted_values_tree),color = "red") + geom_point(aes(y=predicted_values_lm),color = "blue") + geom_abline(aes(intercept=0,slope=1),cex =2) + ylab("Predicted Values")
                                                                                                                                                          
```

When we look at the predicted value points (blue is from lm model and red is from tree model) There is a obvious fit with the x=y line which shows us our predictions fit well to the real price data.
